Formal Grammars as Functors and Formal Grammars as Types in Non-commutative Linear-Non-Linear Type Theory

### 2024-01-24 paper/paper.tex
NFA-tuple to grammar

\subsection{Finite State Machines}
A finite state machine may be expressed via a particular syntactic form for grammars. A \emph{nondeterministic finite automaton} (NFA) is defined as a
mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.}
fixed point in the following manner,

\[
  NFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle Q \rangle \\
        Y_{1} = \langle Q \rangle \\
        \vdots \\
        Y_{e} = I
        \end{pmatrix}~\textbf{in}~Y_{s}
\]
where
\[
  Q ::= c \otimes Y ~|~ Y ~|~ Q \oplus Q
\]
and $Y \in \{ Y_{s}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
definitions in form $Q$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
denotes an $\epsilon$-transition. Without loss of generality, $Y_{s}$ is reserved for a single distinguished start
state, and $Y_{e}$ is reserved for a single distinguished accepting state. A
disjunction\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.}
of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

Let $\hat{Y}_{k}$ refer to the syntactic term on the right hand side of the definition of $Y_{k}$ inside of an NFA $N$.
We say that $N$ is a \emph{deterministic finite automaton} (DFA) if there are no $\epsilon$-transitions and for every
definition $\hat{Y}_{k}$ the $c_{j}$ are distinct guards for each produced state.

% \steven{Is this the right notion of determinism? Perhaps, it is is better to define determinism as something like: there is at most one derivation for all $j, k$, for all $a \in \Sigma$ we have $x : Y_{k}, y : a \vdash z : Y_{j}$.}

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system. First, let us describe a version of Thompson's construction \cite{?} where we show that any regular expression can be recognized by some NFA\@. Moreover, we will show that this NFA is strongly equivalent to the original grammar.

@@

\begin{theorem}
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
\end{theorem}

@@

\[
  NFA(c) =
  \mu \begin{pmatrix}
        Y_{s} = c \otimes Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
\[
  NFA(I) =
  \mu \begin{pmatrix}
        Y_{s} = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]

@@

We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
the disjunction of two NFAs. That is, assuming that $g \cong NFA(g)$ and $g' \cong NFA(g')$, we define an NFA that
recognizes $g \oplus g'$. Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.

\[
  NFA(g \oplus g') =
  \mu \begin{pmatrix}
        Y_{s} =  X_{s}' \oplus Z_{s}' \\
        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
        X'_{e} = Y_{e} \\
        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
        Z_{e}' = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}~\textbf{in}~Y_{s}
\]
where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$. Likewise define
$\delta_{Z}$.
In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
$X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has
an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states
$Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect
of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$
we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
folding is not taken as primitive, using Beki\'c's theorem we can define a family of multifold terms to serve as
elimination forms for our admissible multiple fixed points. That is, rules of the form,

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n})~\textbf{in}~X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

are admissible, where the substitution $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$ is given by $\gamma$. In this instance, we
wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
return a term of type $g \oplus g'$. The structure of such a derivation is given below,

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
This finshes the construction of a parse transformer $NFA(g \oplus g') \to g \oplus g'$.
The map in the other direction is given the derivation,

>>>>

\subsection{Finite State Machines}
A \emph{nondeterministic finite automaton} (NFA) is a finite state machine where transitions are labeled with characters from a fixed alphabet $\Sigma$. These are often represented formally as a 5-tuple $(Q, \Sigma, \delta, q_{0}, F)$,

\begin{itemize}
  \item $Q$ a finite set of states
  \item $\Sigma$ a fixed, finite alphabet
  \item $\delta : Q \times \Sigma \to \mathcal{P}(Q)$ a labelled transition function
  \item $q_{0}$ a start state
  \item $F$ a set of accepting states
\end{itemize}

Intuitively, this can be thought of like a directed graph with nodes in $Q$ with an edge $q \overset{c}{\to} q'$ whenever $q' \in \delta(q, c)$. Note that transitions in an NFA may be labelled with the empty string $\epsilon$.

As a gramamar in our type theory, an NFA may be encoded as a
mutual\footnote{For convenience we refer to a mutual fixed point here, even though we
  only take single fixed points as primitive. As in Beki\'c's theorem, we may decompose our mutual fixed points into a series of single fixed points.}
fixed point in the following manner:

An NFA has finitely many states, so we may number them $Q = \{q_{1}, \dots, q_{N}\}$. For each $q_{j}$ we introduce
a mutually recursive term $Y_{j}$ to represent the states of the NFA within this grammar presentation. We represent the transition function $\delta$ in the right hand side of each $Y_{j}$ definition. For character $c \in \Sigma \cup \{\epsilon\}$ and state $q_{j} \in Q$, we add a summand --- as a coproduct\footnote{There is an admissible notion of coproduct making use of the primitive dependent pairing --- i.e. $A \oplus B := \LinPiTy b \Bool {\textbf{if}~b~A~B}$. We will utilize this without much comment to represent the usual notion of alternation of grammars.} --- to the defintion of $Y_{j}$ for every state in the image of the transition function,

\[
Y_{j} = \bigoplus_{c \in \Sigma \cup \{\epsilon\}} \bigoplus_{q_{i} \in \delta(q_{j}, c)} c \otimes Y_{i}
\]

For convenience, refer to syntactic term on the right hand side of $Y_{j}$ as $\hat{Y}_{j}$. Further, we rename the start state to be represented by the term $Y_{s}$. Additionally, and without loss of generality, we assume that there is a single accepting state $Y_{e}$ in our encoding. This assupmtion is justified through the use of $\epsilon$-transitions. For each $q_{j} \in F$, we add an $\epsilon$-transition from $Y_{j}$ to $Y_{e}$.

The above translation totals to give us a syntactic form for NFAs as grammars,

\[
  NFA ::= \mu \begin{pmatrix}
        Y_{s} = \langle \hat{Y} \rangle \\
        Y_{j} = \langle \hat{Y} \rangle, j \not \in \{s, e\} \\
        Y_{e} = I
        \end{pmatrix}.Y_{s}
\]
where
\[
  \hat{Y} ::= c \otimes Y ~|~ Y ~|~ \hat{Y} \oplus \hat{Y}
\]

and each $Y \in \{ Y_{s}, Y_{e}, Y_{1}, \dots, Y_{N} \}$. Intuitively, we read each $Y_{j}$ as the label for a state, and the
definitions of form $\hat{Y}$ describe the transitions between states --- $c \otimes Y$ denotes a labeled transition, and $Y$
denotes an $\epsilon$-transition. A
disjunction of transitions in an NFA grammar corresponds to multiple outgoing edges of a given state in an automaton.

Above we give a translation from the tuple presentation of an NFA to a fixed point grammar in our type theory. We don't provide any argument that this translation is done correctly; however, we hope that the construction is straightforward and lightweight enough to demonstrate our formal system as a natural and effective syntax for reasoning about both grammars and automata.

With this setting for finite automata, we can now internalize some classical theorems inside of our formal system.

@@

First, let us describe a version of Thompson's construction \cite{?} where we construct an NFA that recognizes a given regular expression. Moreover, we will show that this NFA is strongly equivalent to the original grammar.

This construction

\begin{theorem}[Thompson]
  \label{thm:thompson}
  For $g$ a regular grammar $g$, there is an NFA $N$ that recognizes the same language as $g$. Further, $N$ and $g$ are isomorphic.
\end{theorem}

@@

\[
  NFA(c) =
  \mu \begin{pmatrix}
        Y_{s} = c \otimes Y_{e} \\
        Y_{e} = I
  \end{pmatrix}.Y_{s}
\]
\[
  NFA(I) =
  \mu \begin{pmatrix}
        Y_{s} = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}.Y_{s}
\]

@@

We can now inductively define some NFA combinators and show that they preserve grammar equivalence. First begin with
the disjunction of two NFAs. For regular grammars $g$ and $g'$, assume that $g \cong NFA(g)$ and $g' \cong NFA(g')$, and we will define an NFA that
recognizes $g \oplus g'$.

Let the states in $NFA(g)$ be represented by $X$'s and the states in $NFA(g')$ be represented
by $Z$'s. Further let $\hat{X}_{k}$ refer to the syntactic term on the right hand side of the definition of $X_{k}$
inside of $NFA(g)$. Likewise define each $\hat{Z}_{k}$.

\[
  NFA(g \oplus g') =
  \mu \begin{pmatrix}
        Y_{s} =  X_{s}' \oplus Z_{s}' \\
        X_{j}' = \simulsubst {\hat{X}_{j}} {\delta_{X}}, j \neq e \\
        X'_{e} = Y_{e} \\
        Z_{j}' = \simulsubst {\hat{Z}_{j}} {\delta_{Z}}, j \neq e \\
        Z_{e}' = Y_{e} \\
        Y_{e} = I
  \end{pmatrix}.Y_{s}
\]
where the substitution $\{X_{s}'/X_{s}, X_{e}'/X_{e}, X_{1}'/X_{1}, \dots \}$ is given by $\delta_{X}$ --- and $\delta_{Z}$ constructed similarly.

In the construction of $NFA(g \oplus g')$ we replicate each defintion $X_{j}$ in $NFA(g)$ with an analogous definition
$X_{j}'$ inside of $NFA(g \oplus g')$, except we replace $X_{k}$ with $X_{k}'$ in each production. Lastly, $X_{e}'$ has
an $\epsilon$-transition to the accepting state $Y_{e}$. Likewise, we build analogous defintions for the states
$Z_{j}$ in $NFA(g')$, then we flatten all of these defintions into one mutally recursive list. This has the effect
of disjunctively tying together $NFA(g)$ and $NFA(g')$ into one automaton.

To show the strong equivalence of $NFA(g \oplus g')$ and $g \oplus g'$, we will demonstrate an isomorphism between them
with two proof terms in our logic that mutually invert each other. To define a map $NFA(g \oplus g') \to g \oplus g'$
we will use a multi-fold over all the mutually recursive defintions of $NFA(g \oplus g')$. Note, even though multiple
fixed points are not taken as primitive, using Beki\'c's theorem we can define admissible multiple fixed points types. Below we give rules to form $\mathsf{cons}$ and $\mathsf{mfold}$ terms to serve as the introduction and elimination rules, respectively, for multi-term fixed points,

\[
  \inferrule
  {\Gamma ; \Delta \vdash e : \simulsubst {A_{k}} {\sigma}}
  {\Gamma ; \Delta \vdash \mathsf{cons}~e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{k}}
\]
where $\sigma$ is the substitution that unrolls the mutually recursive definition one level by replacing each $X_{k}$ with the fixed point,
\steven{I'm not sure this is right. TODO sanity check it with a small example}
\begin{align*}
  \sigma = \{ & \mu(X_{1} = A_{1} \dots, X_{n} = A_{n}) . X_{1} / X_{1} , \dots, \\
  & \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}) . X_{n} / X_{n} \}
\end{align*}

\[
  \inferrule{\Gamma ; \Delta \vdash e : \mu(X_{1} = A_{1}, \dots, X_{n} = A_{n}).X_{k} \\
             \Gamma ; x_{1} : \simulsubst {A_{1}}{\gamma} \vdash e_{1} : B_{1} \\
             \dots \\
             \Gamma ; x_{n} : \simulsubst {A_{n}}{\gamma} \vdash e_{n} : B_{n}
  }{\Gamma; \Delta \vdash \mathsf{mfold}(x_{1}.e_{1}, \dots, x_{n}.e_{n})(e) : B_{k}}
\]

where the substitution $\gamma$ represents $\{B_{1}/X_{1},\dots,B_{n}/X_{n} \}$.

When defining a derivation from $NFA(g \oplus g')$ to $g \oplus g'$, we
wish to simultaneously fold over the mutually recursive definitions comprising $NFA(g \oplus g')$. We may think of this
as an eliminator that replaces each state inside of $NFA(g \oplus g')$ with a tail-recursive call that will ultimately
return a term of type $g \oplus g'$. The structure of such a derivation is given below,

@@

where $\phi$ and $\psi$ are the witnesses to the isomorphisms $g \cong NFA(g)$ and $g' \cong NFA(g')$, respectively.
This finshes the construction of a parse transformer from $NFA(g \oplus g')$ to $g \oplus g'$.
The map in the other direction is given the derivation,
